Adaptive global optimization using graphics accelerators\thanks{This study was supported by the Russian Science Foundation, project No.\,16-11-10150.}

### 2020-05-27 2020/RSD_GPU/RSD2020.tex
12 pages again

Thus, calculating the optimization criterion can be implemented on GPU, and the role of the optimization algorithm (running on CPU) consists in the effective selection of points for conducting parallel trials. This scheme of working with the accelerator is fully consistent with the work of the parallel global search algorithm developed at the Lobachevsky State University of Nizhni Novgorod and presented in a series of publications \cite{Barkalov2014,Barkalov2020,Barkalov2019,Grishagin2016,Grishagin2015,Grishagin2018,globalizerSystem}.

@@

In this paper, we will use an approach based on the idea of reducing dimensionality using the Peano space-filling curve \(y(x)\), which continuously and unambiguously maps a segment of the real axis \([0,1]\) onto an \(n\)-dimensional cube
\begin{equation}
\label{cube}
\lbrace y\in R^N:-2^{-1}\leqslant y_i\leqslant 2^{-1},1\leqslant i\leqslant N\rbrace=\{y(x):0\leqslant x\leqslant 1\}.
\end{equation}

The questions of numerical construction of approximations to Piano curve (evolvents) and the corresponding theory are discussed in detail in \cite{Sergeyev2013,Strongin2000}. Using evolvents \(y(x)\) reduces the multidimensional problem (\ref{task}) to a one-dimensional problem
\begin{displaymath}
\label{oneDimTask}
\varphi(y^*)=\varphi(y(x^*))=\min\{\varphi(y(x)):x\in [0,1]\}.
\end{displaymath}

@@

In relation to global optimization methods, an operation that can be efficiently implemented on GPU is the parallel calculation of many values of the objective function at once. Naturally, this requires implementing a procedure for calculating the value of a function on GPU. Data transfers from CPU to GPU will be minimal: one need only transfer the coordinates of the trial points to GPU, and get back the function values at these points. Functions that determine the processing of trial results in accordance with the algorithm, and require work with large amounts of accumulated search information, can be effectively implemented on CPU.

@@

   \begin{figure}[h!]
    \centering
		\includegraphics[width=0.9\textwidth]{GPU_Algorithm.pdf}
		\caption{Scheme of information exchanges in the GPU algorithm}\label{fig:GPU_Algorithm}
\end{figure}

@@

   \begin{figure}[h!]
    \centering
		\includegraphics[width=0.7\textwidth]{Parallel_Computing.pdf}
		\caption{Diagram of parallel computing on a cluster}\label{fig:Parallel_Computing}
\end{figure}

@@

Note that well-known test problems from the field of multidimensional global optimization are characterized by a short time of calculating the values of the objective function. Therefore, in order to simulate the computational complexity inherent in applied optimization problems \cite{Modorskii2016}, the calculation of the objective function in all experiments was complicated by additional calculations that do not change the form of the function and the location of its minima (summing the segment of the Taylor series). 
In the experiments carried out the average time for calculating the function values was 0.01 seconds.

@@

\begin{table}
	\caption{Average number of iterations $k_{av}$}\label{table:average_iters}
	\center
	\begin{tabular}{ccccc}
		\hline\noalign{\smallskip}
  	 $N$ & Problem class & DIRECT & DIRECT\textit{l} & GSA \\
		\noalign{\smallskip} \hline \noalign{\smallskip}
			4 &	Simple& $>$47282(4) &	18983 &	11953 \\
	      & Hard &	$>$95708(7) &	68754 &	25263 \\
		\noalign{\smallskip}
			5	& Simple &	$>$16057(1) &	16758 &	15920 \\
				& Hard &	$>$217215(16) &	$>$269064(4) & $>$148342(4) \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table}

@@

\begin{table}
                \caption{Speedup on CPU}
                \label{table:average_MMAS_CPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
																	 & \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                                   & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=2$&    6,6  & 3,2   &   & 2,1  & 6,6  &   & 3,2 & 1,5 &   & 0,7 & 2,1 \\
                               $p=4$&    21,5 &  10,0 &   & 6,9  & 19,2 &   & 5,2 & 2,4 &   & 0,9 & 2,3 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

\begin{table}
                \caption{Speedup on one GPU}
                \label{table:average_AS_GPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
																& \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    5,1  & 3,9  &   & 1,2 	 & 9,3   &   & 2,2	& 1,9	     &   & 0,5	& 4,0  \\
                               $p=256$ &    19,9 & 15,0 &   & 11,9 	 & 39,6  &   & 8,3	& 6,9	     &   & 3,4	& 11,0 \\
							   $p=1024$&    15,6 & 52,9 &   & 22,1	 & 105,7 &   & 5,2	& 20,0      &   & 2,2	& 10,2 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

\begin{table}
                \caption{Speedup on two GPUs}
                \label{table:average_MMAS_GPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
                & \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    11,3	 & 6,8	 &   & 8,2	 & 19,8  &   & 2,4	 & 1,6	 &   & 1,3	 & 3,2 \\
                               $p=256$ &    39,0	 & 31,5 &   & 25,4	 & 57,1  &   & 8,0	 & 7,0	 &   & 2,3	 & 2,4 \\
							   $p=1024$&    128,4	 & 83,5 &   & 98,4	 & 267,9 &   & 20,4  & 11,0  &   & 2,8	 & 5,2 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

@@

\begin{table}
                \caption{Speedup on six GPUs}
                \label{table:GKLS6_average_MMAS_GPU}
                \center
                \begin{tabular}{lccc}
                               %\hline\noalign{\smallskip}
                &  Iteration speedup & &   Time speedup \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    30,8	 & & 1,9	  \\
                               $p=256$ &    92,7	 & & 1,5  \\
							   $p=1024$&    597,0	 & & 2,5  \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

>>>>

Thus, calculating the optimization criterion can be implemented on GPU, and the role of the optimization algorithm (running on CPU) consists in the selection of points for conducting parallel trials. This scheme of working with the accelerator is fully consistent with the work of the parallel global search algorithm developed at the Lobachevsky State University of Nizhni Novgorod and presented in a series of papers \cite{Barkalov2014,Barkalov2020,Barkalov2019,Grishagin2016,Grishagin2015,Grishagin2018,globalizerSystem}.

@@

In this work we will use an approach based on the idea of reducing dimensionality using the Peano space-filling curve \(y(x)\), which continuously and unambiguously maps a segment of the real axis \([0,1]\) onto an \(n\)-dimensional cube
\begin{equation}
\label{cube}
\lbrace y\in R^N:-2^{-1}\leqslant y_i\leqslant 2^{-1},1\leqslant i\leqslant N\rbrace=\{y(x):0\leqslant x\leqslant 1\}.
\end{equation}

The questions of numerical construction of approximations to Peano curve (\textit{evolvents}) and the corresponding theory are discussed in detail in \cite{Sergeyev2013,Strongin2000}. Using evolvents \(y(x)\) reduces the multidimensional problem (\ref{task}) to a one-dimensional problem
\begin{displaymath}
\label{oneDimTask}
\varphi(y^*)=\varphi(y(x^*))=\min\{\varphi(y(x)):x\in [0,1]\}.
\end{displaymath}

@@

In relation to global optimization methods, an operation that can be efficiently implemented on GPU is the parallel calculation of many values of the objective function at once. Naturally, this requires implementing a procedure for calculating the value of a function on GPU. Data transfers from CPU to GPU will be minimal: one need only transfer the coordinates of the trial points to GPU, and get back the function values at these points. Functions that determine the processing of trial results in accordance with the algorithm, and require work with large amounts of accumulated search information, can be efficiently implemented on CPU.

@@

   \begin{figure}[t]
    \centering
		\includegraphics[width=0.75\textwidth]{GPU_Algorithm.pdf}
		\caption{Scheme of information exchanges in the GPU algorithm}\label{fig:GPU_Algorithm}
\end{figure}

@@

   \begin{figure}[t]
    \centering
		\includegraphics[width=0.65\textwidth]{Parallel_Computing.pdf}
		\caption{Diagram of parallel computing on a cluster}\label{fig:Parallel_Computing}
\end{figure}

@@

Note that well-known test problems from the field of multidimensional global optimization are characterized by a short time of calculating the values of the objective function. Therefore, in order to simulate the computational complexity inherent in applied optimization problems \cite{Modorskii2016}, the calculation of the objective function in all experiments was complicated by additional calculations that do not change the form of the function and the location of its minima (summing the segment of the Taylor series). 
In the experiments carried out the average time for calculating the function values was 0.01 seconds, which exceeds the latency of the network or the data transfer time to the GPU.

@@

\begin{table}[t]
	\caption{Average number of iterations $k_{av}$}\label{table:average_iters}
	\center
	\begin{tabular}{ccccc}
		\hline\noalign{\smallskip}
  	 $N$ & Problem class & DIRECT & DIRECT\textit{l} & GSA \\
		\noalign{\smallskip} \hline \noalign{\smallskip}
			4 &	Simple& $>$47282(4) &	18983 &	11953 \\
	      & Hard &	$>$95708(7) &	68754 &	25263 \\
		\noalign{\smallskip}
			5	& Simple &	$>$16057(1) &	16758 &	15920 \\
				& Hard &	$>$217215(16) &	$>$269064(4) & $>$148342(4) \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table}

@@

\begin{table}[t]
                \caption{Speedup on CPU}
                \label{table:average_MMAS_CPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
																	 & \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                                   & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=2$&    6,6  & 3,2   &   & 2,1  & 6,6  &   & 3,2 & 1,5 &   & 0,7 & 2,1 \\
                               $p=4$&    21,5 &  10,0 &   & 6,9  & 19,2 &   & 5,2 & 2,4 &   & 0,9 & 2,3 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

\begin{table}[t]
                \caption{Speedup on one GPU}
                \label{table:average_AS_GPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
																& \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    5,1  & 3,9  &   & 1,2 	 & 9,3   &   & 2,2	& 1,9	     &   & 0,5	& 4,0  \\
                               $p=256$ &    19,9 & 15,0 &   & 11,9 	 & 39,6  &   & 8,3	& 6,9	     &   & 3,4	& 11,0 \\
							   $p=1024$&    15,6 & 52,9 &   & 22,1	 & 105,7 &   & 5,2	& 20,0      &   & 2,2	& 10,2 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

\begin{table}[t]
                \caption{Speedup on two GPUs}
                \label{table:average_MMAS_GPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
                & \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    11,3	 & 6,8	 &   & 8,2	 & 19,8  &   & 2,4	 & 1,6	 &   & 1,3	 & 3,2 \\
                               $p=256$ &    39,0	 & 31,5 &   & 25,4	 & 57,1  &   & 8,0	 & 7,0	 &   & 2,3	 & 2,4 \\
							   $p=1024$&    128,4	 & 83,5 &   & 98,4	 & 267,9 &   & 20,4  & 11,0  &   & 2,8	 & 5,2 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

@@

\begin{table}[t]
                \caption{Speedup on six GPUs}
                \label{table:GKLS6_average_MMAS_GPU}
                \center
                \begin{tabular}{lccc}
                               %\hline\noalign{\smallskip}
                &  Iteration speedup & &   Time speedup \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    30,8	 & & 1,9	  \\
                               $p=256$ &    92,7	 & & 1,5  \\
							   $p=1024$&    597,0	 & & 2,5  \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}
